Practical Parametricity for GADTs

### 2020-09-23 20-gadts/inf-rules.tex
added functorial actions of set and rel interps of types

{\color{blue} Need to check that the extra condition in the interp of
  Lans is preserved for the last clause to be well-defined.}

@@

\begin{dfn}\label{def:rel-sem}
The {\em relational interpretation}
$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega}$ is defined by
\begin{align*}
  \oCPORsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \oCPORsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho &= \{\eta
  : \lambda \ol{K}.\,\oCPORsem{\Gamma; \Phi \vdash
    F}\rho[\ol{\Phi := K}] \Rightarrow \lambda \ol{K}. \,\oCPORsem{
    \Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
  &=
  \{(\eta_1,\eta_2) \in \oCPOsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G} (\pi_1 \rho) \times \oCPOsem{ 
    \Gamma;\emptyset
    \vdash \Nat^\Phi \,F\,G} (\pi_2
  \rho)~|~\\ 
  & \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}.\\
  & \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
  \oCPORsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
  (\rho\phi)\ol{\oCPORsem{\Gamma;\Phi \vdash 
    F}\rho}\\
  \oCPORsem{\Gamma;\Phi \vdash F+G}\rho &=
  \oCPORsem{\Gamma;\Phi \vdash F}\rho +
  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\
  \oCPORsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \oCPORsem{\Gamma;\Phi \vdash F}\rho \times
  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\  
   \oCPORsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho
  &= (\mu T_{H,\rho})\ol{\oCPORsem{\Gamma;\Phi,\ol\gamma \vdash
     G}\rho}\\
  \text{where }	T_{H,\rho}
    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
  \text{and } T^\oCPOR_{H,\rho}\,F
    &= \lambda \ol{R}. \oCPORsem{
      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
    F][\ol{\alpha := R}]\\
  \text{and } T^\oCPOR_{H,\rho}\,\delta
    &= \lambda \ol{R}. \oCPORsem{
      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
    \delta][\ol{\alpha := \id_{\ol{R}}}] \\
  \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
    F)\ol{A}}\rho &= \{ (t_1, t_2) : \oCPOsem{\Gamma;\Phi \vdash
    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\pi_1 \rho \,\times\,
  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
    F)\ol{A}}\pi_2 \rho\\
  &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma;
    \ol{\alpha} \vdash K}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
  \oCPOsem{\Gamma; \ol{\alpha} \vdash K}\pi_2\rho[\ol{\alpha := Z}],\\
  & \hspace{0.2in} f :
  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to
  \oCPORsem{\Gamma; \Phi \vdash A}\rho\\
  &\hspace{0.2in}\mbox{such that } \iota_{Z,\pi_1 f}t_1' = t_1,\,
  \kappa_{Z,\pi_2 f}t_2' = t_2,\\
  &  \hspace*{0.2in}\mbox{and } (t_1',t_2') :
  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]\,\}
\end{align*}
\end{dfn}
\noindent
{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
  $\iota_{Z,h}$ is the injection into
\[\colim{Z \,: \,\oCPO_0,\; h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
      K}\pi_1\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
      A}\pi_1\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
    F}\pi_1\rho[\ol{\alpha := Z}]}\]
  and 
  $\kappa_{Z,h}$ is the injection into
\[\colim{Z\, :\, \oCPO_0, \;h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
      K}\pi_2\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
    A}\pi_2\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
    F}\pi_2\rho[\ol{\alpha := Z}]}\]}

@@

\begin{dfn}\label{def:rel-sem-funcs}
Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and
$\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash
  F}$ on the morphism $f$ is given exactly as in
Definition~\ref{def:set-sem-funcs}, except that all interpretations
are $\oCPO$ relational interpretations and all occurrences of
$T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.

{\color{blue} This is not true for the relational interps of
  Lans. Define explicitly.}

To see that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined we must show that, for
every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any
$\oCPO$ relation transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\,
F \to T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers
for every $\oCPO$ relation transformer $F$ and every morphism $f :
\rho \to \rho'$ in $\oCPORenv$. This is an immediate consequence of
the following Lemma.

>>>>

We need to check that the functorial action on $\Lan$ types is well-defined.
That amounts to showing that the extra condition in the interpretation of
$\Lan$ types is preserved by the functorial action, so that,
if $t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho$,
then $\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$.
By definition, if $t : \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho$,
then $(t, t) \in \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \Eq_{\rho}$,
i.e., there exist
$\ol{Z : \oCPO_0}$,
$t_1 : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$,
$t_2 : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]$,
and $\ol{g : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\Eq_{\rho}[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\Eq_{\rho}}$
such that
$\iota_{\ol{Z}, \ol{\pi_1 g}}t_1 = t$,
$\kappa_{\ol{Z}, \ol{\pi_2 g}}t_2 = t$,
and
$(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.
To prove that
$\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
: \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'$,
let $\ol{Z' = Z}$, $t'_i = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_{Z}}] t_i$ for $i = 1, 2$,
and $\ol{g' = \oCPORsem{\Gamma; \Phi \vdash A}f \circ g}$.
We thus have to show that
$\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$,
$\kappa'_{\ol{Z'}, \ol{\pi_2 g'}}t'_2 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$,
and
$(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha := \Eq_{Z'}}]$.
That $(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho'}[\ol{\alpha := \Eq_{Z'}}]$
follows from the fact that
$(t_1',t_2') = \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} Eq_{f} [\ol{\alpha := \id_{\Eq_{Z}}}] (t_1, t_2)$
and $(t_1,t_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\Eq_{\rho}[\ol{\alpha := \Eq_Z}]$.
That $\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$
is because
\begin{array}{rl}
\iota'_{\ol{Z'}, \ol{\pi_1 g'}}t'_1
&= \iota'_{\ol{Z'}, \ol{\pi_1 g'}} (\oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_{Z}}] t_1) \\
&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f (\iota_{\ol{Z}, \ol{\pi_1 g}} t_1) \\
&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t
\end{array}
where the second equality is justified by the commutativity of the diagram
\begin{equation}\label{cd:lan-comm}
\begin{tikzcd}[row sep = huge]
\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
  \ar[r, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
  \ar[d, "{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}" description]
&\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
  \ar[d, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}}" description] \\
\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
  \ar[r, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
  \ar[d, equal]
& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
  \ar[d, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}f}}" description] \\
\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
  \ar[r, shift right = 1, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}f \circ g)}}}"']
& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
\end{equation}
%\[
%\begin{tikzcd}
%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := S}]}
%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}f[\ol{\alpha := \id_{\_}}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}\rho}}"]
%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := Z}]
%  \ar[u, "{\iota_{\ol{Z}, \ol{\pi_1 g}}}"]
%  \ar[r, "{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \id_Z}]}"]
%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
%\end{tikzcd}
%\]
%and
%\[
%\begin{tikzcd}
%\colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}}
%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]}
%  \ar[r, bend left = 10, "{(\Lan_{\ol{\oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \_}]}} \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \_}]) \ol{\oCPORsem{\Gamma; \Phi \vdash A}f}}"]
%& \colim{\ol{S}, \ol{h : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho'[\ol{\alpha := S}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho'}}
%{\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := S}]} \\
%\oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
%  \ar[u, "{\iota''_{\ol{Z}, \ol{\pi_1 g}}}"]
%  \ar[r, equal]
%& \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := Z}]
%  \ar[u, "{\iota'_{\ol{Z'}, \ol{\pi_1 (\oCPORsem{\Gamma; \Phi \vdash A}f \circ g)}}}"]
%\end{tikzcd}
%\]
Analogously, we show that $\kappa_{Z,\pi_2 g'}t'_2 = \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}f t$.

@@

\begin{dfn}\label{def:rel-sem}
The {\em relational interpretation}
$\oCPORsem{\cdot} : \F \to [\oCPORenv, \oCPOR]_{\omega}$ is defined by
\begin{align*}
  \oCPORsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \oCPORsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \oCPORsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho &= \{\eta
  : \lambda \ol{K}.\,\oCPORsem{\Gamma; \Phi \vdash
    F}\rho[\ol{\Phi := K}] \Rightarrow \lambda \ol{K}. \,\oCPORsem{
    \Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
  &=
  \{(\eta_1,\eta_2) \in \oCPOsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G} (\pi_1 \rho) \times \oCPOsem{ 
    \Gamma;\emptyset
    \vdash \Nat^\Phi \,F\,G} (\pi_2
  \rho)~|~\\ 
  & \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : \oCPORT_{k}}.\\
  & \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
  (\oCPORsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi :=
      K}])^{\oCPORsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
  \oCPORsem{\Gamma;\Phi \vdash \phi \ol{F}}\rho &=
  (\rho\phi)\ol{\oCPORsem{\Gamma;\Phi \vdash 
    F}\rho}\\
  \oCPORsem{\Gamma;\Phi \vdash F+G}\rho &=
  \oCPORsem{\Gamma;\Phi \vdash F}\rho +
  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\
  \oCPORsem{\Gamma;\Phi \vdash F\times G}\rho &=
  \oCPORsem{\Gamma;\Phi \vdash F}\rho \times
  \oCPORsem{\Gamma;\Phi \vdash G}\rho\\  
   \oCPORsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho
  &= (\mu T_{H,\rho})\ol{\oCPORsem{\Gamma;\Phi,\ol\gamma \vdash
     G}\rho}\\
  \text{where }	T_{H,\rho}
    &= (T^\set_{H,\pi_1\rho}, T^\set_{H,\pi_2\rho}, T^\rel_{H,\rho}) \\
  \text{and } T^\oCPOR_{H,\rho}\,F
    &= \lambda \ol{R}. \oCPORsem{
      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\rho[\phi :=
    F][\ol{\alpha := R}]\\
  \text{and } T^\oCPOR_{H,\rho}\,\delta
    &= \lambda \ol{R}. \oCPORsem{
      \Gamma;\ol\gamma,\phi,\ol{\alpha} \vdash H}\id_\rho[\phi :=
    \delta][\ol{\alpha := \id_{\ol{R}}}] \\
  \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
    F)\ol{A}}\rho &= \{ (t_1, t_2) : \oCPOsem{\Gamma;\Phi \vdash
    (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\pi_1 \rho \,\times\,
  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}}
    F)\ol{A}}\pi_2 \rho\\
  &\hspace{0.2in}|~\exists\, \ol{Z : \oCPO_0}, t_1' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_1\rho[\ol{\alpha := Z}],\, t_2' :
  \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_2\rho[\ol{\alpha := Z}],\\
  & \hspace{0.2in} \ol{f :
  \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to
  \oCPORsem{\Gamma; \Phi \vdash A}\rho}\\
  &\hspace{0.2in}\mbox{such that } \iota_{\ol{Z},\ol{\pi_1 f}}t_1' = t_1,\,
  \kappa_{\ol{Z}, \ol{\pi_2 f}}t_2' = t_2,\\
  &  \hspace*{0.2in}\mbox{and } (t_1',t_2') :
  \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]\,\}
\end{align*}
\end{dfn}
\noindent
{\color{blue} In the final clause of Definition~\ref{def:rel-sem},
  $\iota_{Z,h}$ is the injection into
\[\colim{Z \,: \,\oCPO_0,\; h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
      K}\pi_1\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
      A}\pi_1\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
    F}\pi_1\rho[\ol{\alpha := Z}]}\]
  and 
  $\kappa_{Z,h}$ is the injection into
\[\colim{Z\, :\, \oCPO_0, \;h \,:\, \oCPOsem{\Gamma; \ol{\alpha} \vdash
      K}\pi_2\rho[\ol{\alpha := Z}] \to \oCPOsem{\Gamma; \Phi \vdash
    A}\pi_2\rho}{\oCPOsem{\Gamma; \ol{\alpha} \vdash
    F}\pi_2\rho[\ol{\alpha := Z}]}\]}

@@

\begin{dfn}\label{def:rel-sem-funcs}
Let $f: \rho \to \rho'$ for $\oCPO$ relation environments $\rho$ and
$\rho'$ (so that $\rho|_\tvars = \rho'|_\tvars$). The action
$\oCPORsem{\Gamma;\Phi \vdash F}f$ of $\oCPORsem{\Gamma;\Phi \vdash F}$
on the morphism $f$ for all types but $\Lan$ is given exactly as in
Definition~\ref{def:set-sem-funcs}, except that all interpretations
are $\oCPO$ relational interpretations and all occurrences of
$T^\oCPO_{H,f}$ are replaced by $T_{H,f}$.
In the $\Lan$ case, instead, define
\[
\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} f
%: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho
%\to \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
%\]
%as
%\[
= (
\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f,
\oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f
)
\]
\end{dfn}

We need to show that the functorial action for the $\Lan$ case is well-defined, i.e.,
that for every $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$
we have that
\[
(
\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f t_1,
\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f t_2
)
: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
\]
From $(t_1, t_2) : \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}$
we have, by definition, that there exist
$\ol{Z : \oCPO_0}$,
$t_1' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_1\rho[\ol{\alpha := Z}]$,
$t_2' : \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F}\pi_2\rho[\ol{\alpha := Z}]$,
$\ol{g : \oCPORsem{\Gamma; \ol{\alpha} \vdash K}\rho[\ol{\alpha := \Eq_Z}] \to \oCPORsem{\Gamma; \Phi \vdash A}\rho}$
such that
$\iota_{\ol{Z},\ol{\pi_1 g}}t_1' = t_1$,
$\kappa_{\ol{Z}, \ol{\pi_2 g}}t_2' = t_2$,
and
$(t_1',t_2') : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho[\ol{\alpha := \Eq_Z}]$.
To show that
\[
(
\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_1 f t_1,
\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \pi_2 f t_2
)
: \oCPORsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} \rho'
\]
let
$\ol{S = Z : \oCPO_0}$,
$s_1 = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f) [\ol{\alpha := Z}] t'_1$,
$s_2 = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_2 f) [\ol{\alpha := Z}] t'_2$,
and
$\ol{h = \oCPORsem{\Gamma; \Phi \vdash A} f \circ g}$.
The first thing we need to show is that $(s_1, s_2) : \oCPORsem{\Gamma; \Phi, \ol\alpha \vdash F}\rho'[\ol{\alpha := \Eq_S}]$,
and that is because $(s_1, s_2) = \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} f [\ol{\alpha := \Eq_Z}] (t'_1, t'_2)$.
The second thing we need to show is that $\iota'_{\ol{S},\ol{\pi_1 h}} s_1' = s_1$ and $\kappa'_{\ol{S},\ol{\pi_2 h}} s_2' = s_2$.
That is because
\[
\begin{array}{rl}
\iota'_{\ol{S},\ol{\pi_1 h}} s_1'
&= \iota'_{\ol{S},\ol{\oCPOsem{\Gamma; \Phi \vdash A} \pi_1 f \circ \pi_1 g}} ( \oCPOsem{\Gamma; \Phi, \ol\alpha \vdash F} (\pi_1 f) [\ol{\alpha := Z}] t'_1 ) \\
&= (\oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) \circ \iota_{\ol{S},\ol{\pi_1 g}}) t'_1 \\
&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) (\iota_{\ol{S},\ol{\pi_1 g}} t'_1) \\
&= \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}} (\pi_1 f) t_1 \\
&= s_1
\end{array}
\]
where the second equality is by the commutativity of diagram~\ref{cd:lan-comm}.
Analogously, we show that $\kappa'_{\ol{S},\ol{\pi_2 h}} s_2' = s_2$.

To see that the functors given by Definitions~\ref{def:rel-sem}
and~\ref{def:rel-sem-funcs} are well-defined we must also show that, for
every $H$, $T_{H,\rho}\,F$ is an $\oCPO$ relation transformer for any
$\oCPO$ relation transformer $F$, and that $T_{H,f}\, F : T_{H,\rho}\,
F \to T_{H,\rho'}\, F$ is a morphism of $\oCPO$ relation transformers
for every $\oCPO$ relation transformer $F$ and every morphism $f :
\rho \to \rho'$ in $\oCPORenv$. This is an immediate consequence of
the following Lemma.
